---
title: "CSphere"
author: "Michayla Moore"
date: "2024-08-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Install required packages if you don't have them
#install.packages("GEOquery")
#install.packages("Seurat")  # or install BiocManager::install("SingleCellExperiment")
#BiocManager::install("GEOquery")

install.packages('devtools')
devtools::install_github('immunogenomics/presto')

# Load the packages
library(Matrix)
library(Seurat)
library(GEOquery)
library(ggplot2)
library(tidyverse)
library(reshape2)
library(dplyr)
library(patchwork)
library(writexl)
library(tibble)
library(RColorBrewer)
```

# Extract Raw Data From GEO
```{r}
# Load the GEO dataset
gse <- getGEO("GSE149699", GSEMatrix = TRUE)
```

```{r}
# Raw data
# Download supplementary files for GSE149699
getGEOSuppFiles("GSE149699")

# Unzip files (if needed)
#unzip("GSE149699_RAW.tar", exdir = "GSE149699_files")
untar("GSE149699_RAW.tar", exdir = "GSE149699_files")
```

```{r}
# List all the zipped files in the directory
zip_files <- list.files("GSE149699_files", pattern = "\\.gz$", full.names = TRUE)
# List all .gz files
gz_files <- list.files("GSE149699_files", pattern = "\\.gz$", full.names = TRUE)

for (gz_file in gz_files) {
  sample_name <- tools::file_path_sans_ext(basename(gz_file))
  target_dir <- paste0("GSE149699_files/", sample_name)
  
  # Create a directory for the sample if it doesn't exist
  if (!dir.exists(target_dir)) {
    dir.create(target_dir, recursive = TRUE)
  }
  
  # Define the output file path
  output_file <- paste0(target_dir, "/", tools::file_path_sans_ext(basename(gz_file)))
  
  # Unzip the .gz file into the target directory
  tryCatch({
    gunzip(gz_file, destname = output_file, overwrite = TRUE)
  }, error = function(e) {
    message("Error unzipping file: ", gz_file, "\n", e)
  })
}
```

```{r}
# Get all file names
file_names <- list.files("GSE149699_files", full.names = TRUE)

# Extract unique sample identifiers
sample_ids <- unique(gsub("_.*", "", basename(file_names)))

# Print sample IDs to verify
print(sample_ids)
```

# Define Data Path and Name for Files and Create Seurat Object
```{r, warning=FALSE}
# Define file paths for each sample - change as needed
EC_barcodes <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509138_EC_barcodes.tsv/GSM4509138_EC_barcodes.tsv"
EC_features <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509138_EC_features.tsv/GSM4509138_EC_features.tsv"
EC_matrix <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509138_EC_matrix.mtx/GSM4509138_EC_matrix.mtx"

CF_barcodes <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509139_CF_barcodes.tsv/GSM4509139_CF_barcodes.tsv"
CF_features <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509139_CF_features.tsv/GSM4509139_CF_features.tsv"
CF_matrix <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509139_CF_matrix.mtx/GSM4509139_CF_matrix.mtx"

SMC_barcodes <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509140_SMC_barcodes.tsv/GSM4509140_SMC_barcodes.tsv"
SMC_features <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509140_SMC_features.tsv/GSM4509140_SMC_features.tsv"
SMC_matrix <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509140_SMC_matrix.mtx/GSM4509140_SMC_matrix.mtx"

Ad_barcodes <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509141_adult_barcodes.tsv/GSM4509141_adult_barcodes.tsv"
Ad_features <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509141_adult_features.tsv/GSM4509141_adult_features.tsv"
Ad_matrix <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509141_adult_matrix.mtx/GSM4509141_adult_matrix.mtx"

Neo_barcodes <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509142_neonatal_barcodes.tsv/GSM4509142_neonatal_barcodes.tsv"
Neo_features <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509142_neonatal_features.tsv/GSM4509142_neonatal_features.tsv"
Neo_matrix <- "\\\\mhirfs/vary_lab/Michayla/CSPhere/GSE149699_files/GSM4509142_neonatal_matrix.mtx/GSM4509142_neonatal_matrix.mtx"

# Load the data for sample 1
barcodesEC <- read.table(EC_barcodes, header = FALSE, stringsAsFactors = FALSE)[[1]]
featuresEC <- read.table(EC_features, header = FALSE, stringsAsFactors = FALSE)
matrixEC <- readMM(EC_matrix)

# Create Seurat object for sample 1
seurat_sampleEC <- CreateSeuratObject(counts = matrixEC, assay = "RNA", project = "EC")
rownames(seurat_sampleEC) <- featuresEC$V2
colnames(seurat_sampleEC) <- barcodesEC

# Load the data for sample 2
barcodesCF <- read.table(CF_barcodes, header = FALSE, stringsAsFactors = FALSE)[[1]]
featuresCF <- read.table(CF_features, header = FALSE, stringsAsFactors = FALSE)
matrixCF <- readMM(CF_matrix)

# Create Seurat object for sample 2
seurat_sampleCF <- CreateSeuratObject(counts = matrixCF, assay = "RNA", project = "CF")
rownames(seurat_sampleCF) <- featuresCF$V2
colnames(seurat_sampleCF) <- barcodesCF

# Load the data for sample 3
barcodesSMC <- read.table(SMC_barcodes, header = FALSE, stringsAsFactors = FALSE)[[1]]
featuresSMC <- read.table(SMC_features, header = FALSE, stringsAsFactors = FALSE)
matrixSMC <- readMM(SMC_matrix)

# Create Seurat object for sample 3
seurat_sampleSMC <- CreateSeuratObject(counts = matrixSMC, assay = "RNA", project = "SMC")
rownames(seurat_sampleSMC) <- featuresSMC$V2
colnames(seurat_sampleSMC) <- barcodesSMC

# Load the data for sample 4
barcodesAd <- read.table(Ad_barcodes, header = FALSE, stringsAsFactors = FALSE)[[1]]
featuresAd <- read.table(Ad_features, header = FALSE, stringsAsFactors = FALSE)
matrixAd <- readMM(Ad_matrix)

# Create Seurat object for samplev 4
seurat_sampleAd <- CreateSeuratObject(counts = matrixAd, assay = "RNA", project = "Ad")
rownames(seurat_sampleAd) <- featuresAd$V2
colnames(seurat_sampleAd) <- barcodesAd

# Load the data for sample 5
barcodesNeo <- read.table(Neo_barcodes, header = FALSE, stringsAsFactors = FALSE)[[1]]
featuresNeo <- read.table(Neo_features, header = FALSE, stringsAsFactors = FALSE)
matrixNeo <- readMM(Neo_matrix)

# Create Seurat object for sample 5
seurat_sampleNeo <- CreateSeuratObject(counts = matrixNeo, assay = "RNA", project = "Neo")
rownames(seurat_sampleNeo) <- featuresNeo$V2
colnames(seurat_sampleNeo) <- barcodesNeo
```


# Name Orig.Ident Files and Remove Duplicates
```{r, include=FALSE}
seurat_sampleEC$orig.ident <- "EC"
seurat_sampleCF$orig.ident <- "CF"
seurat_sampleSMC$orig.ident <- "SMC"
seurat_sampleAd$orig.ident <- "AdultCDC"
seurat_sampleNeo$orig.ident <- "NeonatalCDC"

# Identify duplicate gene names
duplicated_genes <- rownames(seurat_sampleEC)[duplicated(rownames(seurat_sampleEC))]

# Print duplicated genes to confirm
cat("Duplicate genes found:\n", unique(duplicated_genes), "\n")

# Identify rows to remove (those with duplicate names)
dup_indices <- which(duplicated(rownames(seurat_sampleEC)) | duplicated(rownames(seurat_sampleEC), fromLast = TRUE))

# Remove duplicate rows
seurat_sampleEC_clean <- seurat_sampleEC[-dup_indices, ]

# Verify removal
remaining_duplicates <- rownames(seurat_sampleEC_clean)[duplicated(rownames(seurat_sampleEC_clean))]
cat("Remaining duplicate genes (if any):\n", unique(remaining_duplicates), "\n")

# Replace the original object with the cleaned one
seurat_sampleEC <- seurat_sampleEC_clean

#CF
# Identify duplicate gene names
duplicated_genesCF <- rownames(seurat_sampleCF)[duplicated(rownames(seurat_sampleCF))]

# Print duplicated genes to confirm
cat("Duplicate genes found:\n", unique(duplicated_genesCF), "\n")

# Identify rows to remove (those with duplicate names)
dup_indicesCF <- which(duplicated(rownames(seurat_sampleCF)) | duplicated(rownames(seurat_sampleCF), fromLast = TRUE))

# Remove duplicate rows
seurat_sampleCF_clean <- seurat_sampleCF[-dup_indicesCF, ]

# Verify removal
remaining_duplicatesCF <- rownames(seurat_sampleCF_clean)[duplicated(rownames(seurat_sampleCF_clean))]
cat("Remaining duplicate genes (if any):\n", unique(remaining_duplicatesCF), "\n")

# Replace the original object with the cleaned one
seurat_sampleCF <- seurat_sampleCF_clean

# Function to remove duplicate gene names from a Seurat object
clean <- function(seurat_object) {
  # Identify duplicate gene names
  duplicated_genes <- rownames(seurat_object)[duplicated(rownames(seurat_object))]
  
  # Print duplicated genes to confirm
  cat("Duplicate genes found:\n", unique(duplicated_genes), "\n")
  
  # Identify rows to remove (those with duplicate names)
  dup_indices <- which(duplicated(rownames(seurat_object)) | duplicated(rownames(seurat_object), fromLast = TRUE))
  
  # Remove duplicate rows
  seurat_object_clean <- seurat_object[-dup_indices, ]
  
  # Verify removal
  remaining_duplicates <- rownames(seurat_object_clean)[duplicated(rownames(seurat_object_clean))]
  cat("Remaining duplicate genes (if any):\n", unique(remaining_duplicates), "\n")
  
  # Replace the original object with the cleaned one
  seurat_object <- seurat_object_clean
  
  # Normalize the data
  #seurat_object <- NormalizeData(seurat_object)
  
  return(seurat_object)
}

# Apply the function to each Seurat object
seurat_sampleCF <- clean(seurat_sampleCF)
seurat_sampleSMC <- clean(seurat_sampleSMC)
seurat_sampleAd <- clean(seurat_sampleAd)
seurat_sampleNeo <- clean(seurat_sampleNeo)
```

# Filter Low Quality and Mitochondrial Genes 
```{r}
# Function to calculate mitochondrial percentage
calculate_mito <- function(seurat_obj) {
  seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
  return(seurat_obj)
}

# Apply function to all datasets
seurat_sampleEC <- calculate_mito(seurat_sampleEC)
seurat_sampleCF <- calculate_mito(seurat_sampleCF)
seurat_sampleSMC <- calculate_mito(seurat_sampleSMC)
seurat_sampleAd <- calculate_mito(seurat_sampleAd)
seurat_sampleNeo <- calculate_mito(seurat_sampleNeo)

# Function to plot histograms
plot_histograms <- function(seurat_obj, title) {
  p1 <- ggplot(seurat_obj@meta.data, aes(x = nFeature_RNA)) + 
    geom_histogram(binwidth = 100, fill = "blue", alpha = 0.6) + 
    ggtitle(paste(title, "- Number of Features")) + 
    theme_minimal()
  
  p2 <- ggplot(seurat_obj@meta.data, aes(x = nCount_RNA)) + 
    geom_histogram(binwidth = 500, fill = "green", alpha = 0.6) + 
    ggtitle(paste(title, "- Number of Counts")) + 
    theme_minimal()
  
  p3 <- ggplot(seurat_obj@meta.data, aes(x = percent.mt)) + 
    geom_histogram(binwidth = 1, fill = "red", alpha = 0.6) + 
    ggtitle(paste(title, "- Mitochondrial Percentage")) + 
    theme_minimal()
  
  return(p1 + p2 + p3)
}

# Generate histograms for each dataset
plot_histograms(seurat_sampleEC, "EC")
plot_histograms(seurat_sampleCF, "CF")
plot_histograms(seurat_sampleSMC, "SMC")
plot_histograms(seurat_sampleAd, "Ad")
plot_histograms(seurat_sampleNeo, "Neo")
```

# Apply Filter for Low Quality and Mitochondrial Genes  
```{r}
# Define function to filter out low-quality cells
filter_cells <- function(seurat_obj) {
  # Filter based on mitochondrial percentage (<20%) and nFeature_RNA (200-2500)
  seurat_obj <- subset(seurat_obj, subset = percent.mt < 20 & nFeature_RNA > 200 & nCount_RNA < 150000) # if wanted "& nFeature_RNA < 2500"
  return(seurat_obj)
}

# Apply filtering function to each dataset
seurat_sampleCF_filtered <- filter_cells(seurat_sampleCF)
seurat_sampleSMC_filtered <- filter_cells(seurat_sampleSMC)
seurat_sampleAd_filtered <- filter_cells(seurat_sampleAd)
seurat_sampleNeo_filtered <- filter_cells(seurat_sampleNeo)
seurat_sampleEC_filtered <- filter_cells(seurat_sampleEC)

# Function to plot histograms after filtering
plot_histograms_after_filtering <- function(seurat_obj, title) {
  p1 <- ggplot(seurat_obj@meta.data, aes(x = nFeature_RNA)) + 
    geom_histogram(binwidth = 100, fill = "blue", alpha = 0.6) + 
    ggtitle(paste(title, "- Number of Features")) + 
    theme_minimal()
  
  p2 <- ggplot(seurat_obj@meta.data, aes(x = nCount_RNA)) + 
    geom_histogram(binwidth = 500, fill = "green", alpha = 0.6) + 
    ggtitle(paste(title, "- Number of Counts")) + 
    theme_minimal()
  
  p3 <- ggplot(seurat_obj@meta.data, aes(x = percent.mt)) + 
    geom_histogram(binwidth = 1, fill = "red", alpha = 0.6) + 
    ggtitle(paste(title, "- Mito %")) + 
    theme_minimal()
  
  return(p1 + p2 + p3)
}

# Generate histograms for filtered datasets
plot_histograms_after_filtering(seurat_sampleEC_filtered, "EC")
plot_histograms_after_filtering(seurat_sampleCF_filtered, "CF")
plot_histograms_after_filtering(seurat_sampleSMC_filtered, "SMC")
plot_histograms_after_filtering(seurat_sampleAd_filtered, "Ad")
plot_histograms_after_filtering(seurat_sampleNeo_filtered, "Neo")
```


# Check Filtering
```{r}
# Function to create boxplots before and after filtering
create_boxplot <- function(seurat_obj_before, seurat_obj_after, feature, feature_name) {
  data_before <- data.frame(
    Sample = rep("Before Filtering", length(seurat_obj_before@meta.data[[feature]])),
    Value = seurat_obj_before@meta.data[[feature]],
    filtered = "Before Filtering"
  )
  
  data_after <- data.frame(
    Sample = rep("After Filtering", length(seurat_obj_after@meta.data[[feature]])),
    Value = seurat_obj_after@meta.data[[feature]],
    filtered = "After Filtering"
  )
  
  combined_data <- rbind(data_before, data_after)
  
  # Plotting the boxplot
  ggplot(combined_data, aes(x = filtered, y = Value, fill = filtered)) +
    geom_boxplot() +
    facet_wrap(~Sample) +
    labs(title = feature_name, y = feature, x = "") +
    #theme_minimal() +
    theme(axis.text.x = element_blank()) +
    scale_fill_manual(values = c("Before Filtering" = "grey", "After Filtering" = "blue"))
}

# Create boxplots for nCount, nFeature, and mito for each sample (before and after filtering)
create_boxplot(seurat_sampleCF, seurat_sampleCF_filtered, "nCount_RNA", "Distribution of nCount (CF)")
create_boxplot(seurat_sampleSMC, seurat_sampleSMC_filtered, "nCount_RNA", "Distribution of nCount (SMC)")
create_boxplot(seurat_sampleAd, seurat_sampleAd_filtered, "nCount_RNA", "Distribution of nCount (Ad)")
create_boxplot(seurat_sampleNeo, seurat_sampleNeo_filtered, "nCount_RNA", "Distribution of nCount (Neo)")

create_boxplot(seurat_sampleCF, seurat_sampleCF_filtered, "nFeature_RNA", "Distribution of nFeature (CF)")
create_boxplot(seurat_sampleSMC, seurat_sampleSMC_filtered, "nFeature_RNA", "Distribution of nFeature (SMC)")
create_boxplot(seurat_sampleAd, seurat_sampleAd_filtered, "nFeature_RNA", "Distribution of nFeature (Ad)")
create_boxplot(seurat_sampleNeo, seurat_sampleNeo_filtered, "nFeature_RNA", "Distribution of nFeature (Neo)")
```

# Normalize, Feature, Scale, and Run PCA on the Filtered Data
```{r}
# Normalize
seurat_sampleCF_norm <- NormalizeData(seurat_sampleCF_filtered)
seurat_sampleSMC_norm <- NormalizeData(seurat_sampleSMC_filtered)
seurat_sampleAd_norm <- NormalizeData(seurat_sampleAd_filtered)
seurat_sampleNeo_norm <- NormalizeData(seurat_sampleNeo_filtered)
seurat_sampleEC_norm <- NormalizeData(seurat_sampleEC_filtered)

# Feature
seurat_sampleCF_norm <- FindVariableFeatures(seurat_sampleCF_norm)
seurat_sampleSMC_norm <- FindVariableFeatures(seurat_sampleSMC_norm)
seurat_sampleAd_norm <- FindVariableFeatures(seurat_sampleAd_norm)
seurat_sampleNeo_norm <- FindVariableFeatures(seurat_sampleNeo_norm)
seurat_sampleEC_norm <- FindVariableFeatures(seurat_sampleEC_norm)

# Scale
seurat_sampleCF_norm <- ScaleData(seurat_sampleCF_norm)
seurat_sampleSMC_norm <- ScaleData(seurat_sampleSMC_norm)
seurat_sampleAd_norm <- ScaleData(seurat_sampleAd_norm)
seurat_sampleNeo_norm <- ScaleData(seurat_sampleNeo_norm)
seurat_sampleEC_norm <- ScaleData(seurat_sampleEC_norm)

# Run PCA
seurat_sampleCF_norm <- RunPCA(seurat_sampleCF_norm)
seurat_sampleSMC_norm <- RunPCA(seurat_sampleSMC_norm)
seurat_sampleAd_norm <- RunPCA(seurat_sampleAd_norm)
seurat_sampleNeo_norm <- RunPCA(seurat_sampleNeo_norm)
seurat_sampleEC_norm <- RunPCA(seurat_sampleEC_norm)
```

## Note: Raw Counts (@assays$RNA@counts) contain the raw expression counts for each gene across cells. These are the original values obtained from sequencing and have not been transformed in any way.

## Normalized Data (@assays$RNA@data) is a transformation of the raw counts. In Seurat, this usually involves log-normalization, where each gene's counts are divided by the total counts for that cell (to account for differences in sequencing depth), followed by log transformation. This results in a matrix that is usually in a transformed scale.

# Check Normalization
```{r}
# Violin plot for a specific gene before and after normalization
VlnPlot(seurat_sampleCF_filtered, features = "ACTB", layer = "counts") + ggtitle("Unnormalized (Raw Counts) - CF") +
  scale_fill_manual(values = "grey")
VlnPlot(seurat_sampleCF_norm, features = "ACTB", layer = "data") + ggtitle("Normalized - CF") +
  scale_fill_manual(values = "red")

# Boxplot for ACTB before and after normalization
df_non_norm <- data.frame(Expression = c(GetAssayData(seurat_sampleCF_filtered, slot = "counts")["ACTB", ]),
                          Type = "Non-Normalized")
df_norm <- data.frame(Expression = c(GetAssayData(seurat_sampleCF_norm, slot = "data")["ACTB", ]),
                      Type = "Normalized")

df_combined <- rbind(df_non_norm, df_norm)
ggplot(df_combined, aes(x = Type, y = Expression, fill = Type)) +
  geom_boxplot() +
  ggtitle("Boxplot: ACTB Expression Before and After Normalization") +
  scale_fill_manual(values = c("grey", "red")) +
  theme_minimal()

# Calculate average expression
avg_expr_non_norm <- rowMeans(GetAssayData(seurat_sampleCF_filtered, slot = "counts"))
avg_expr_norm <- rowMeans(GetAssayData(seurat_sampleCF_norm, slot = "data"))

# Reshape the data into a long format
df_avg_expr <- data.frame(
  Gene = rep(rownames(seurat_sampleCF_filtered), 2),  # Repeat Gene names for both conditions
  Average_Expression = c(avg_expr_non_norm, avg_expr_norm),
  Type = rep(c("Non-Normalized", "Normalized"), each = length(avg_expr_non_norm))
)

# Create the box plot
ggplot(df_avg_expr, aes(x = Type, y = Average_Expression, fill = Type)) +
  geom_boxplot() +
  ggtitle("Comparison of Average Gene Expression (All Data)") +
  theme_minimal() +
  ylab("Average Expression") +
  xlab("Normalization Type") +
  scale_fill_manual(values = c("grey", "red"))  
  #scale_y_continuous(limits = c(-50, 50))  # Set y-axis limits
```


# Find Features (Most Variable Genes) and Anchors (Batch Effects)
## When using Seurat's FindIntegrationAnchors() and IntegrateData() functions, it is generally considered best practice to provide unnormalized Seurat objects as input because these functions perform their own normalization steps as part of the integration process
```{r}
# List datasets (change clean to norm 3/6/25)
seurat_objects <- list(seurat_sampleAd_norm, seurat_sampleCF_norm, seurat_sampleEC_norm, seurat_sampleNeo_norm, seurat_sampleSMC_norm)

# Ident integration features
# The function identifies genes that show variation across the datasets, which are likely to contribute to the biological signal during # the integration process. This helps reduce noise and ensures that the integration process focuses on the relevant aspects of the data.
features <- SelectIntegrationFeatures(object.list = seurat_objects)

# returns only the genes that are present in all Seurat objects.
common_features <- Reduce(intersect, lapply(seurat_objects, function(x) rownames(GetAssayData(x, assay = "RNA", layer = "counts"))))

# By subsetting each dataset to only include the common features, you ensure that the integration step is performed on the exact same 
# set of genes across all datasets. 
seurat_objects <- lapply(seurat_objects, function(x) {
  x <- subset(x, features = common_features)
  return(x)
})

# Align
# Anchors are pairs of cells from different datasets that are biologically similar and can be used as reference points 
# for aligning the datasets.
anchors <- FindIntegrationAnchors(object.list = seurat_objects, anchor.features = features)
```

# Integrate Data (Make seurat_combined)
```{r}
# integrate data
seurat_combined <- IntegrateData(anchorset = anchors)
```

# Histograms of Seurat_combined
```{r}
# Check available assays and their slots
# Assays(seurat_combined)

# Plot histogram for number of features
hist(seurat_combined$nFeature_RNA, main = "Number of Features per Cell", xlab = "Number of Features", col = "blue", breaks = 50)

# Plot histogram for number of Counts
hist(seurat_combined$nCount_RNA , main = "Number of Counts per Cell", xlab = "Number of Counts", col = "grey", breaks = 50)

# Plot histogram for number of Mito
hist(seurat_combined$percent.mt , main = "Number of Mito per Cell", xlab = "Number of Mito", col = "red", breaks = 50)

# Add metadata
meta <- seurat_combined@meta.data
```


# Combine Counts from All Groups Into New "seurat_combined"
```{r}
# Assuming matrices
combined_counts <- cbind(seurat_combined@assays$RNA$counts.Ad, 
                         seurat_combined@assays$RNA$counts.CF, 
                         seurat_combined@assays$RNA$counts.SMC, 
                         seurat_combined@assays$RNA$counts.Neo, 
                         seurat_combined@assays$RNA$counts.EC)  

# Create Seurat object with combined counts
seurat_combined <- CreateSeuratObject(counts = combined_counts)
```

# Feature, Scale, and PCA
```{r}
# Normalize 
seurat_combined <- NormalizeData(seurat_combined)

# Continue
seurat_combined <- FindVariableFeatures(seurat_combined)
seurat_combined <- ScaleData(seurat_combined)
seurat_combined <- RunPCA(seurat_combined)

print(seurat_combined[["pca"]])

# Recompute PCA
# seurat_combined <- RunPCA(seurat_combined, npcs = 30)

# Print PCA summary
# print(seurat_combined[["pca"]])
```

# Find "dims" (Optimal Number of Principal Components)
```{r}
# Common dims is 10-20
# Where does graph taper off?
ElbowPlot(seurat_combined)
```


# Find Neighbors and Clusters
```{r}
# This calculates which cells are closest to each other in the lower-dimensional space
seurat_combined <- FindNeighbors(seurat_combined, dims = 1:14)

# This clustering algorithm (Louvain or Leiden by default) to identify distinct groups of cells based on their gene expression profiles.
# Default is 0.8
seurat_combined <- FindClusters(seurat_combined)
```

# Run UMAP
```{r}
# Add or update metadata in seurat_combined
seurat_combined <- AddMetaData(seurat_combined, metadata = meta)

# View metadata
head(seurat_combined@meta.data)

#Run UMAP (Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization)
seurat_combined <- RunUMAP(seurat_combined, dims = 1:14)
DimPlot(seurat_combined, reduction = "umap", group.by = "orig.ident")
```

# Customize UMAP
```{r, warning=FALSE}
# Define custom colors for your specific cell types including AdultCDC and NeonatalCDC
custom_colors <- c("CF" = "darkgreen", 
                   "EC" = "purple", 
                   "SMC" = "darkorange", 
                   "AdultCDC" = "blue", 
                   "NeonatalCDC" = "red")

# Create UMAP plot with custom colors for specific cell types and CDC groups
umap_plot <- DimPlot(seurat_combined, 
                      reduction = "umap", 
                      group.by = "orig.ident", 
                      cols = custom_colors) +  
          theme(legend.position = c(0, 1),  
          legend.justification = c("left", "top"))  

umap_plot

# Save the plot as a PNG image with dimensions 5x5 inches
ggsave("umap_plot.png", plot = umap_plot, width = 5, height = 5, dpi = 1200)
```


# Basic Summary Data
```{r}
# All Genes
all_features <- rownames(seurat_combined[["RNA"]])
all_features_df <- as.data.frame(all_features)
head(all_features_df)

table(seurat_combined$orig.ident)       # Number of cells per group/sample
summary(seurat_combined@meta.data)  # Summary of all metadata fields


summary(seurat_combined$nCount_RNA)  # RNA counts per cell
summary(seurat_combined$nFeature_RNA)  # Number of genes per cell
summary(seurat_combined$percent.mt)  # Mitochondrial gene percentage

VlnPlot(seurat_combined, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"))
```


# How Many Cells for Analysis? 
# What is Average Expression of Entire Dataset?
```{r}
# Access metadata from the Seurat object
metadata <- seurat_combined@meta.data

# Count the number of cells per orig.ident
cell_counts <- table(metadata$orig.ident)

# Display the result
print(cell_counts)

# Calculate average expression levels by cluster
avg_expression <- AggregateExpression(seurat_combined, group.by = "orig.ident")

# Extract the expression data (use "counts" or "data" depending on your needs)
expression_data <- GetAssayData(seurat_combined, assay = "RNA", layer = "data")  # or "counts"

# Calculate the average expression for the whole dataset
average_expression <- mean(expression_data)

# Output the result
cat("Average expression for the entire dataset:", average_expression, "\n")
```


# Graph Specific Gene (FeaturePlot)
```{r}
# Extract UMAP coordinates
umap_coords <- Embeddings(seurat_combined, reduction = "umap")

# Extract gene expression data and metadata
gene_data <- FetchData(seurat_combined, vars = c("CXCL6", "orig.ident"))

# Combine UMAP coordinates with gene expression data
umap_data <- cbind(umap_coords, gene_data)

#umap_data <- umap_data %>%
 # rename(CD73 = NT5E)

# Reshape the data to long format
umap_long <- melt(umap_data, id.vars = c("umap_1", "umap_2", "orig.ident"), 
                  measure.vars = c("CXCL6"),
                  variable.name = "gene", value.name = "expression")

# Plot
umap_plot2 <- ggplot(umap_long, aes(x = umap_1, y = umap_2, color = expression)) +
  geom_point() +
  facet_grid(gene ~ orig.ident) +  # Facet by gene and cell type
  scale_color_gradient(low = "lightgrey", high = "blue") +
  theme_minimal() +
  ggtitle("Gene Expression") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),  # Center and enlarge title
    axis.title = element_text(size = 14, face = "bold"),  # Axis labels
    axis.text = element_text(size = 14),  # Tick labels
    strip.text = element_text(size = 14, face = "italic"),  # Facet labels
    legend.title = element_text(size = 14, face = "bold"),  # Legend title
    legend.text = element_text(size = 14)  # Legend labels
  )

# Save the plot as a PNG image with dimensions 5x5 inches
ggsave("gene_expression_umap.png", plot = umap_plot2, width = 10, height = 5, dpi = 1200)

umap_plot2
```


# Plot Multiple Genes in One UMap
```{r}
# Extract UMAP coordinates
umap_coords <- Embeddings(seurat_combined, reduction = "umap")

# Extract gene expression data and metadata
gene_data <- FetchData(seurat_combined, vars = c("ENG", "ACVRL1", "SOST", "IGFBP3", "ISLR", "orig.ident"))

# Combine UMAP coordinates with gene expression data
umap_data <- cbind(umap_coords, gene_data)

umap_data <- umap_data %>%
  rename(CD105 = ENG)

# Reshape the data to long format
umap_long <- melt(umap_data, id.vars = c("umap_1", "umap_2", "orig.ident"), 
                  measure.vars = c("CD105", "ACVRL1", "SOST", "IGFBP3", "ISLR"),
                  variable.name = "gene", value.name = "expression")

# Use ggplot to create custom plots with different colors for each gene
ggplot(umap_long, aes(x = umap_1, y = umap_2, color = expression)) +
  geom_point() +
  facet_grid(gene ~ orig.ident) +  # Facet by gene and cell type
  scale_color_gradient(low = "lightgrey", high = "blue") +
  theme_minimal() +
  ggtitle("Gene Expression") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    strip.text = element_text(size = 10, face = "italic")
  )
```

# Statistical Comparisons of Gene Across Cell Types (MANUAL)
## Wilcoxon Rank Sum Test with Bonferroni Correction
```{r}
seurat_combined <- SetIdent(seurat_combined, value = "orig.ident")

# Compare AdultCDC to CF
markers_adultcdc_vs_cf <- FindMarkers(seurat_combined, ident.1 = "AdultCDC", ident.2 = "CF",
                                      logfc.threshold = 0, min.pct = 0, only.pos = FALSE)

# Compare AdultCDC to EC
markers_adultcdc_vs_ec <- FindMarkers(seurat_combined, ident.1 = "AdultCDC", ident.2 = "EC",
                                      logfc.threshold = 0, min.pct = 0, only.pos = FALSE)

# Compare AdultCDC to SMC
markers_adultcdc_vs_smc <- FindMarkers(seurat_combined, ident.1 = "AdultCDC", ident.2 = "SMC",
                                       logfc.threshold = 0, min.pct = 0, only.pos = FALSE)

# Compare AdultCDC to NeonatalCDC
markers_adultcdc_vs_neonatalcdc <- FindMarkers(seurat_combined, ident.1 = "AdultCDC", ident.2 = "NeonatalCDC",
                                               logfc.threshold = 0, min.pct = 0, only.pos = FALSE)

## Markers_adultcdc_vs_neonatalcdc 
markers_neonatalcdc_vs_adultcdc <- FindMarkers(seurat_combined, ident.1 = "NeonatalCDC", ident.2 = "AdultCDC", 
                                               logfc.threshold = 0, min.pct = 0, only.pos = FALSE)
```

# Top Genes (20; sorted by LogFC) for Each Comparison
```{r}
# Find top expressing genes in each comparison
top_genes_adultcdc_vs_cf <- markers_adultcdc_vs_cf %>%
  arrange(desc(avg_log2FC)) %>%
  head(20)

top_genes_adultcdc_vs_ec <- markers_adultcdc_vs_ec %>%
  arrange(desc(avg_log2FC)) %>%
  head(20)

top_genes_adultcdc_vs_smc <- markers_adultcdc_vs_smc %>%
  arrange(desc(avg_log2FC)) %>%
  head(20)

top_genes_adultcdc_vs_neonatalcdc <- markers_adultcdc_vs_neonatalcdc %>%
  arrange(desc(avg_log2FC)) %>%
  head(20)

# Combine top genes from all comparisons
combined_top_genes <- bind_rows(
  top_genes_adultcdc_vs_cf %>% mutate(comparison = "AdultCDC vs CF"),
  top_genes_adultcdc_vs_ec %>% mutate(comparison = "AdultCDC vs EC"),
  top_genes_adultcdc_vs_smc %>% mutate(comparison = "AdultCDC vs SMC"),
  top_genes_adultcdc_vs_neonatalcdc %>% mutate(comparison = "AdultCDC vs NeonatalCDC")
)

# View combined results
head(combined_top_genes)
```

# Statistical Comparisons of Gene Across Cell Types (FUNCTION)
## Wilcoxon Rank Sum Test with Bonferroni Correction
```{r}
# Function to compare AdultCDC with other groups
compare_groups <- function(seurat_combined, group1, group2) {
  markers <- FindMarkers(seurat_combined, ident.1 = group1, ident.2 = group2, logfc.threshold = 0, min.pct = 0, only.pos = FALSE)
  return(markers)
}

# Define the comparisons
comparisons <- c("CF", "EC", "SMC", "NeonatalCDC")

# Initialize an empty list to store results
markers_list <- list()

# Loop through comparisons and store results in the list
for (comparison in comparisons) {
  markers_name <- paste("markers_adultcdc_vs_", comparison, sep = "")
  markers_list[[markers_name]] <- compare_groups(seurat_combined, "AdultCDC", comparison) # change as needed
}
```

# Save Marker Comparisons Results
```{r}
# Make row names as a column called "Gene"
markers_adultcdc_vs_neonatalcdc_save <- rownames_to_column(markers_adultcdc_vs_neonatalcdc, var = "Gene")

# Save the data frame as an Excel file
write_xlsx(markers_adultcdc_vs_neonatalcdc_save, "markers_adultcdc_vs_neonatalcdc.xlsx")
```

# Volcano Plots
```{r, warning=FALSE}
# Define the volcano plot function
volcano <- function(df, title) {
  # Ensure row names are moved to a column
  df <- data.frame(Gene_Name = rownames(df), df, row.names = NULL)

  # Add significance labels
  df$Significance <- "Not Significant"
  df$Significance[df$p_val_adj < 0.000001 & df$avg_log2FC > 1.7] <- "Upregulated"
  df$Significance[df$p_val_adj < 0.000001 & df$avg_log2FC < -1.7] <- "Downregulated"

  # Labels for significant genes
  df$delabel <- NA
  df$delabel[df$Significance != "Not Significant"] <- df$Gene_Name[df$Significance != "Not Significant"]

  # Generate the volcano plot
  p <- ggplot(df, aes(x = avg_log2FC, y = -log10(p_val_adj), color = Significance, label = delabel)) +
    geom_point() + 
    theme_minimal() +
    ggrepel::geom_text_repel(show.legend = FALSE) +
    scale_color_manual(values = c("Upregulated" = "red", "Not Significant" = "grey", "Downregulated" = "blue")) +
    geom_vline(xintercept = c(-1.7, 1.7), col = "black") +
    geom_hline(yintercept = -log10(0.000001), col = "black") +
    labs(color = 'Differentially Expressed',
         x = "Log Fold Change",
         y = "-log10 p-value",
         title = title) +  
    theme(
      axis.title = element_text(size = 16),  
      axis.text = element_text(size = 16),  
      plot.title = element_text(size = 16, hjust=0.5),  
      legend.text = element_text(size = 12), 
      legend.title = element_text(size = 12)
    )

  print(p)  # Print the plot
  
  # Save the plot
  ggsave(paste0("volcano_", gsub(" ", "_", title), ".png"), plot = p, width = 6, height = 4, dpi = 300)
}

# Example usage with multiple datasets
datasets <- list(
  markers_adultcdc_vs_cf = markers_adultcdc_vs_cf, 
  markers_adultcdc_vs_ec = markers_adultcdc_vs_ec,
  markers_adultcdc_vs_smc = markers_adultcdc_vs_smc,
  markers_adultcdc_vs_neonatalcdc = markers_adultcdc_vs_neonatalcdc
)

# Loop through datasets and generate volcano plots
for (name in names(datasets)) {
  volcano(datasets[[name]], paste("Volcano Plot:", name))
}
```


# Graph Stats Results for AdultCDC to All Other Cells Comparison Only (One Gene Only)
```{r}
# Extract logFC for CXCL6 from each comparison
logfc_values <- data.frame(
  Comparison = c("AdultCDC vs CF", "AdultCDC vs EC", "AdultCDC vs SMC", "AdultCDC vs NeonatalCDC"),
  logFC = c(
    markers_adultcdc_vs_cf["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_ec["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_smc["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_neonatalcdc["CXCL6", "avg_log2FC"]
  )
)

# Extract logFC and p-values for CXCL6 from each comparison
logfc_and_pvalues <- data.frame(
  Comparison = c("AdultCDC vs CF", "AdultCDC vs EC", "AdultCDC vs SMC", "AdultCDC vs NeonatalCDC"),
  logFC = c(
    markers_adultcdc_vs_cf["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_ec["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_smc["CXCL6", "avg_log2FC"],
    markers_adultcdc_vs_neonatalcdc["CXCL6", "avg_log2FC"]
  ),
  p_value_adj = c(
    markers_adultcdc_vs_cf["CXCL6", "p_val_adj"],
    markers_adultcdc_vs_ec["CXCL6", "p_val_adj"],
    markers_adultcdc_vs_smc["CXCL6", "p_val_adj"],
    markers_adultcdc_vs_neonatalcdc["CXCL6", "p_val_adj"]
  )
)

# Add significance stars based on p-value
logfc_and_pvalues$significance <- cut(logfc_and_pvalues$p_value,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", "ns"),
  right = FALSE
)

ggplot(logfc_and_pvalues, aes(x = Comparison, y = logFC, fill = Comparison)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = significance), vjust = -0.5, size = 5) +
  theme_minimal() +
  labs(title = "CXCL6",
       x = NULL,
       y = "Log Fold Change") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5,face = "italic")) +
  scale_fill_brewer(palette = "Dark2") +
  ylim(-2, 10)

cxcl6 <- ggplot(logfc_and_pvalues, aes(x = Comparison, y = logFC, fill = Comparison)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = significance), vjust = -0.5, size = 5) +
  theme_minimal() +
  labs(title = "CXCL6",
       x = NULL,
       y = "Log Fold Change") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2") +
  ylim(-2, 10) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "italic"),  # Center and enlarge title
    axis.title = element_text(size = 14, face = "italic"),  # Axis labels
    axis.text = element_text(size = 14),  # Tick labels
    strip.text = element_text(size = 14),  # Facet labels
    legend.title = element_text(size = 14, face = "bold"),  # Legend title
    legend.text = element_text(size = 14)  # Legend labels
  )

cxcl6
```

# Plot Single Analysis Only (i.e. NeoCDC vs AdultCDC)
```{r}
# Extract logFC for "gene" from each comparison
logfc_values <- data.frame(
  Comparison = c("NeonatalCDC vs AdultCDC"),
  logFC = c(
    markers_neonatalcdc_vs_adultcdc["NRDC", "avg_log2FC"]
  )
)

# Extract logFC and p-values for NRDC from each comparison
logfc_and_pvalues <- data.frame(
  Comparison = c("NeonatalCDC vs AdultCDC"),
  logFC = c(
    markers_neonatalcdc_vs_adultcdc["NRDC", "avg_log2FC"]
  ),
  p_value_adj = c(
    markers_neonatalcdc_vs_adultcdc["NRDC", "p_val_adj"]
  )
)

# Add significance stars based on p-value
logfc_and_pvalues$significance <- cut(logfc_and_pvalues$p_value_adj,
    breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "ns"),
    right = FALSE,
    include.lowest = TRUE)

gene1 <- ggplot(logfc_and_pvalues, aes(x = Comparison, y = logFC, fill = Comparison)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = significance), vjust = -0.5, size = 5) +
  theme_minimal() +
  labs(title = "NeonatalCDC vs AdultCDC",
       x="NRDC",
       y = "Log Fold Change") +
  theme(axis.text.x = element_blank()) +
  scale_fill_brewer(palette = "Dark2") +
  ylim(-0.25, 1) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),  # Center and enlarge title
    axis.title = element_text(size = 12, face = "bold"),  # Axis labels
    axis.text = element_text(size = 12),  # Tick labels
    strip.text = element_text(size = 12, face = "italic"),  # Facet labels
    legend.title = element_text(size = 12, face = "bold"),  # Legend title
    legend.text = element_text(size = 12)  # Legend labels
  )

# Convert logFC to FC
logfc_and_pvalues$FC <- 2^logfc_and_pvalues$logFC

# Plot fold change (FC)
gene1_fc <- ggplot(logfc_and_pvalues, aes(x = Comparison, y = FC, fill = Comparison)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = significance), vjust = -0.5, size = 5) +
  theme_minimal() +
  labs(title = "NeonatalCDC vs AdultCDC",
       x="NRDC",
       y = "Fold Change") +
  theme(axis.text.x = element_blank()) +
  scale_fill_brewer(palette = "Spectral") +
  ylim(0, 5) +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),  # Center and enlarge title
    axis.title = element_text(size = 12, face = "bold"),  # Axis labels
    axis.text = element_text(size = 12),  # Tick labels
    strip.text = element_text(size = 12, face = "italic"),  # Facet labels
    legend.title = element_text(size = 12, face = "bold"),  # Legend title
    legend.text = element_text(size = 12)  # Legend labels
  )

gene1_fc
gene1
logfc_and_pvalues

#combined_plot <- gene1 + gene1_fc
#print(combined_plot)
```


# Plot Stats for MULTIPLE GENES
```{r}
# List of genes to analyze
genes <- c("MMP1", "CXCL5", "NT5E", "TFP12", "LRP1", "IGFBP7")

# Create an empty list to store plots
plots <- list()

# Loop through each gene
for (gene in genes) {
  
  # Extract logFC for each gene from each comparison
  logfc_values <- data.frame(
    Comparison = c("AdultCDC vs CF", "AdultCDC vs EC", "AdultCDC vs SMC", "AdultCDC vs NeonatalCDC"),
    logFC = c(
      markers_adultcdc_vs_cf[gene, "avg_log2FC"],
      markers_adultcdc_vs_ec[gene, "avg_log2FC"],
      markers_adultcdc_vs_smc[gene, "avg_log2FC"],
      markers_adultcdc_vs_neonatalcdc[gene, "avg_log2FC"]
    )
  )
  
  # Extract logFC and p-values for each gene from each comparison
  logfc_and_pvalues <- data.frame(
    Comparison = c("AdultCDC vs CF", "AdultCDC vs EC", "AdultCDC vs SMC", "AdultCDC vs NeonatalCDC"),
    logFC = c(
      markers_adultcdc_vs_cf[gene, "avg_log2FC"],
      markers_adultcdc_vs_ec[gene, "avg_log2FC"],
      markers_adultcdc_vs_smc[gene, "avg_log2FC"],
      markers_adultcdc_vs_neonatalcdc[gene, "avg_log2FC"]
    ),
    p_value_adj = c(
      markers_adultcdc_vs_cf[gene, "p_val_adj"],
      markers_adultcdc_vs_ec[gene, "p_val_adj"],
      markers_adultcdc_vs_smc[gene, "p_val_adj"],
      markers_adultcdc_vs_neonatalcdc[gene, "p_val_adj"]
    )
  )
  
  # Add significance stars based on p-value
  logfc_and_pvalues$significance <- cut(logfc_and_pvalues$p_value_adj,
    breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "ns"),
    right = FALSE
  )
  
  # Create the plot for the current gene
  plot <- ggplot(logfc_and_pvalues, aes(x = Comparison, y = logFC, fill = Comparison)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = significance), vjust = -0.5, size = 6.5) +
    theme_minimal() +
    labs(title = paste(gene),
         x = NULL,
         y = "Log Fold Change") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
           plot.title = element_text(hjust = 0.5, size = 12, face = "italic")) + 
    scale_fill_brewer(palette = "Dark2") +
    ylim(-10, 20)
  
  # Add the plot to the list of plots
  plots[[gene]] <- plot
}

# Display all plots
for (gene in genes) {
  print(plots[[gene]])
}

logfc_and_pvalues
```


# Cell Count Analysis for Specific Gene
```{r}
# Fetch expression data for Gene
ISLR_expression <- FetchData(seurat_combined, vars = "ISLR")

# Add the cell type labels (assuming your cell types are stored in the "cell_type" metadata)
ISLR_data <- data.frame(CellType = seurat_combined@meta.data$orig.ident, ISLR = ISLR_expression)

# Check how many cells express NT5E (non-zero counts)
#ISLR_non_zero <- ISLR_data %>%
 # filter(ISLR > 0) %>%
  #group_by(CellType) %>%
  #summarise(count = n())

#print(ISLR_non_zero)

# Summarize expression data by cell type
summary_data <- ISLR_data %>%
  group_by(CellType) %>%
  summarise(mean_expression = mean(ISLR, na.rm = TRUE),
            num_cells = n())

print(summary_data)

# Boxplot of expression by cell type
ggplot(ISLR_data, aes(x = CellType, y = ISLR, col="red")) +
  geom_boxplot() +
  labs(title = "Expression of ISLR by Cell Type", y = "ISLR Expression") +
  theme_minimal()
```


# Heat Maps (Default Pearson)
```{r, warning=FALSE}
# Extract expression data for the genes of interest
expression_data <- FetchData(seurat_combined, vars = c("NT5E", "CXCL6", "CTHRC1", "COL3A1", "HLAB", "CREG1", "COL6A1", "MRC2", "CTHRC1", "COL5A1", "TFP12", "LRP1", "IGFBP7", "ICAM2", "CDH5", "VWF"))

# Compute the correlation matrix
correlation_matrix <- cor(expression_data, use = "pairwise.complete.obs")

# Print the correlation matrix
#print(correlation_matrix)

# Convert correlation matrix to long format for ggplot
correlation_melted <- melt(correlation_matrix)

# heat map correlation heatmap
ggplot(correlation_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Pearson", x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5"))  +
  scale_y_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5")) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3)

# Spearman Check (for non-linear relationship)
# Extract expression data for the genes of interest
expression_data2 <- FetchData(seurat_combined, vars = c("NT5E", "CXCL6", "CTHRC1", "COL3A1", "HLAB", "CREG1", "COL6A1", "MRC2", "CTHRC1", "COL5A1", "TFP12", "LRP1", "IGFBP7", "ICAM2", "CDH5", "VWF"))

# Compute the correlation matrix
correlation_matrix2 <- cor(expression_data, method = "spearman", use = "pairwise.complete.obs")

# Print the correlation matrix
#print(correlation_matrix)

# Convert correlation matrix to long format for ggplot
correlation_melted2 <- melt(correlation_matrix2)

# heat map correlation heatmap
ggplot(correlation_melted2, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Spearman", x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5"))  +
  scale_y_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5")) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3)
```


# Heat Map for Different Cell-Type Markers (Pearson)
```{r}
expression_data <- FetchData(seurat_combined, vars = c("RPL22",
"FAH",
"PPP1R14B",
"RPL27A",
"HNRNPH1",
"AHNAK",
"RAB11B",
"ARCN1",
"GLO1",
"B4GAT1",
"CXCL6", "CXCL5", "THY1", "GATA4",
"ICAM2", "CDH5", "VWF",
"VIM", "FN1", "POSTN"))


# Compute the correlation matrix
correlation_matrix <- cor(expression_data, method = "pearson", use = "pairwise.complete.obs")

# Print the correlation matrix
#print(correlation_matrix)

# Convert correlation matrix to long format for ggplot
correlation_melted <- melt(correlation_matrix)

# heat map correlation heatmap
ggplot(correlation_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "pink", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "THY1" = "CD90"))  +
  scale_y_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "THY1" = "CD90")) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 2)
```

# More Heat Maps
```{r}
# Extract expression data for the genes of interest
expression_data <- FetchData(seurat_combined, vars = c("SOST", "ISLR", "IGFBP3", "ACVRL1", "ENG", "TGFBR2", "ICAM2", "CDH5", "VWF"))

# Compute the correlation matrix
correlation_matrix <- cor(expression_data, use = "pairwise.complete.obs")

# Print the correlation matrix
#print(correlation_matrix)

# Convert correlation matrix to long format for ggplot
correlation_melted <- melt(correlation_matrix)

# heat map correlation heatmap
ggplot(correlation_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "green", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(x = "", y = "", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5"))  +
  scale_y_discrete(labels = c("ACVRL1" = "ACVRL1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "CDH5" = "CDH5")) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3)
```



# Correlation and Adj. P-Values
```{r}
# Example: Compute multiple correlations and adjust p-values
genes <- c("SOST", "CXCL6", "IGFBP3", "ISLR", "CDH5", "GATA4")  # Example gene list
cor_results <- lapply(combn(genes, 2, simplify = FALSE), function(pair) {
  g1 <- FetchData(seurat_combined, vars = pair[1])
  g2 <- FetchData(seurat_combined, vars = pair[2])
  test <- cor.test(g1[,1], g2[,1], method = "pearson")
  return(c(pair[1], pair[2], test$estimate, test$p.value))
})

# Convert to data frame
cor_results_df <- do.call(rbind, cor_results)
colnames(cor_results_df) <- c("Gene1", "Gene2", "Correlation", "P.Value")

# Convert matrix to data frame explicitly
cor_results_df <- as.data.frame(do.call(rbind, cor_results), stringsAsFactors = FALSE)

# Rename columns correctly
colnames(cor_results_df) <- c("Gene1", "Gene2", "Correlation", "P.Value")

# Convert numeric columns from character (if needed)
cor_results_df$Correlation <- as.numeric(cor_results_df$Correlation)
cor_results_df$P.Value <- as.numeric(cor_results_df$P.Value)

# Adjust p-values using Benjamini-Hochberg (FDR)
cor_results_df$Adj.P.Value <- p.adjust(cor_results_df$P.Value, method = "BH")

# View results
print(cor_results_df)
```


# Different Types of Plots (UMAP and Dot Plot)
```{r, warning=FALSE}
# Plot the expression of ACVRL1, CXCL6, ENG, and CD73
FeaturePlot(seurat_combined, features = c("CXCL6", "CTHRC1", "NT5E", "ENG"), ncol = 2) +
  scale_x_discrete(labels = c("CXCL6" = "CXCL6", "NT5E" = "CD73"))

# Create a DotPlot for the genes of interest
DotPlot(seurat_combined, 
        features = c("EGFR", "ERBB2", "ERBB3", "ERBB4", "NRG1", "EGF", "CANX", "PYCARD", "DDX5", "TOMM40"), 
        group.by = "orig.ident") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_x_discrete(
          labels = c("EGFR" = "ERBB1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73")
          )
```

# Linear Regression Per Cell Type
```{r}
# Define genes of interest
gene1 <- "CXCL5"
gene2 <- "CXCL6"

# Extract expression data for each cell type
cell_types <- c("AdultCDC", "CF", "NeonatalCDC")
for (cell_type in cell_types) {
  
  # Subset Seurat object to specific cell type
  sub_data <- subset(seurat_combined, idents = cell_type)
  
  # Extract expression values
  expr_data <- FetchData(sub_data, vars = c(gene1, gene2))
  
  # Perform linear regression
  lm_fit <- lm(expr_data[[gene2]] ~ expr_data[[gene1]])
  summary_fit <- summary(lm_fit)
  
  # Extract R² value and p-value
  r_squared <- round(summary_fit$r.squared, 3)
  p_value <- signif(summary_fit$coefficients[1,4], 3)  # p-value

  # Print summary
  print(paste("Linear Regression for", cell_type))
  print(summary_fit)  # Shows β₁, p-value, R²
  
  # Create the plot with annotation for R² and p-value
  p <- ggplot(expr_data, aes(x = !!sym(gene1), y = !!sym(gene2))) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "blue") +
    theme_minimal() +
    labs(title = paste(cell_type, ":", gene1, "vs", gene2),
         x = gene1, y = gene2) +
    annotate("text", x = min(expr_data[[gene1]], na.rm = TRUE), 
             y = max(expr_data[[gene2]], na.rm = TRUE), 
             label = paste("R² =", r_squared, "\n p =", p_value),
             hjust = -1, vjust = 1, size = 5, color = "black") +
    theme(plot.title = element_text(hjust = 0.5, size = 14))
  
  print(p)
}

boxplot(expr_data$CXCL5, main = "Expression Distribution", col="red")
boxplot(expr_data$CXCL6, main = "Expression Distribution", col="blue")

```

# Linear Regression Across All Cells
```{r}
# Define genes of interest
gene1 <- "CXCL5"
gene2 <- "CXCL6"

# Extract expression data for all cells (no cell type subset)
expr_data <- FetchData(seurat_combined, vars = c(gene1, gene2))

# Perform linear regression
lm_fit <- lm(expr_data[[gene2]] ~ expr_data[[gene1]])
summary_fit <- summary(lm_fit)

# Extract R² value and p-value
r_squared <- round(summary_fit$r.squared, 3)
p_value <- signif(summary_fit$coefficients[2, 4], 3)  # p-value for gene1 coefficient

# Print summary
#print("Linear Regression across all cell types")
#print(summary_fit)  # Shows β₁, p-value, R²

# Create the plot with annotation for R² and p-value
p <- ggplot(expr_data, aes(x = !!sym(gene1), y = !!sym(gene2))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  theme_minimal() +
  labs(title = paste(gene1, "vs", gene2, "Across All Cells"),
       x = gene1, y = gene2) +
  annotate("text", x = min(expr_data[[gene1]], na.rm = TRUE), 
           y = max(expr_data[[gene2]], na.rm = TRUE), 
           label = paste("R² =", r_squared, "\n p =", p_value),
           hjust = -1, vjust = 1, size = 5, color = "black") +
    theme(plot.title = element_text(hjust = 0.5, size = 14))

# Print the plot
print(p)

# Extract expression data for all cells (no cell type subset)
expr_data <- FetchData(seurat_combined, vars = c(gene1, gene2))

# Remove rows where either gene has zero or NA values
expr_data_clean <- expr_data[expr_data[[gene1]] != 0 & expr_data[[gene2]] != 0 & !is.na(expr_data[[gene1]]) & !is.na(expr_data[[gene2]]), ]

# Perform linear regression on the cleaned data
lm_fit <- lm(expr_data_clean[[gene2]] ~ expr_data_clean[[gene1]])
summary_fit <- summary(lm_fit)

# Extract R² value and p-value
r_squared <- round(summary_fit$r.squared, 3)
p_value <- signif(summary_fit$coefficients[2, 4], 3)  # p-value for gene1 coefficient

# Print summary
#print("Linear Regression across all cell types (after removing zeros/NA)")
#print(summary_fit)  # Shows β₁, p-value, R²

# Create the plot with annotation for R² and p-value
p <- ggplot(expr_data_clean, aes(x = !!sym(gene1), y = !!sym(gene2))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  theme_minimal() +
  labs(title = paste(gene1, "vs", gene2, "Across All Cells (Remove Zeros)"),
       x = gene1, y = gene2) +
  annotate("text", x = min(expr_data_clean[[gene1]], na.rm = TRUE), 
           y = max(expr_data_clean[[gene2]], na.rm = TRUE), 
           label = paste("R² =", r_squared, "\n p =", p_value),
           hjust = -1, vjust = 1, size = 5, color = "black") +
    theme(plot.title = element_text(hjust = 0.5, size = 14))

# Print the plot
print(p)
```


# Thresholds for SCRNASeq genes Across Cell Types (i.e. SOST)
```{r}
# Check if Gene is present in the rownames
if ("SOST" %in% rownames(avg_expression)) {
  # Get the average expression in each cluster
  sost_expression_avg <- avg_expression["SOST", ]
  
  # Display the average expression levels
  print(sost_expression_avg)
} else {
  cat("Gene is not found in the average expression data.\n")
}

# Fetch expression data for SOST across all cells
sost_data <- FetchData(seurat_combined, vars = "SOST")

# Add the orig.ident (cell type) as a metadata column
sost_data$orig.ident <- seurat_combined$orig.ident

# Calculate average expression for SOST by cluster
sost_avg_by_cluster <- aggregate(sost_data$SOST, by = list(sost_data$orig.ident), FUN = mean)

# Rename the columns for clarity
colnames(sost_avg_by_cluster) <- c("Cluster", "Avg_SOST_Expression")

# Display the average expression levels
print(sost_avg_by_cluster)
```


# Violin Plots
```{r, warning=FALSE}
features <- c("NRDC", "BMP10", "KIT")
features1 <- c("GDF2", "ACVRL1", "ISLR")
features2 <- c("SOST", "IGFBP3", "ISLR")
features3 <- c("EGFR", "ERBB2", "ERBB4")
features4 <- c("CDH5", "VWF", "PECAM1")
features5 <- c("CXCL6", "ENG", "ACVRL1")
features6 <- c("IGFBP2", "INMT", "SBSPON")
features7 <- c("ACTA2", "CALD1", "TAGLN")
features8 <- c("SOST", "LRP5", "LRP6")

VlnPlot(seurat_combined, features = features) 
VlnPlot(seurat_combined, features = features1)
VlnPlot(seurat_combined, features = features2)
VlnPlot(seurat_combined, features = features3)
VlnPlot(seurat_combined, features = features4)
VlnPlot(seurat_combined, features = features5)
VlnPlot(seurat_combined, features = features6)
VlnPlot(seurat_combined, features = features7)
VlnPlot(seurat_combined, features = features8)
```


# Dot Plots (Change Color, etc.)
```{r}
# Create a DotPlot for the genes of interest
DotPlot(seurat_combined, 
        features = c("CXCL6", "NT5E", "CTHRC1", "ACVRL1"), 
        group.by = "orig.ident") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_x_discrete(
          labels = c("CTHRC1" = "CTHRC1", "CXCL6" = "CXCL6", "ENG" = "CD105", "NT5E" = "CD73", "ACVRL1" = "ALK1")) +
        scale_color_gradient(low = "blue", high = "red")   

DotPlot(seurat_combined, 
        features = c("CXCL6", "NT5E", "CTHRC1", "ACVRL1"), 
        group.by = "orig.ident") +
  theme_minimal()  

DotPlot(seurat_combined, 
        features = c("CXCL6", "NT5E", "CTHRC1", "ACVRL1"), 
        group.by = "orig.ident") +
  scale_size(range = c(4, 4)) +  # Fixed dot size
  scale_color_gradientn(colors = c("blue", "white", "red"))
```

# Co-Expression Analysis (FeaturePlot)
```{r}
FeaturePlot(seurat_combined, features = c("SOST", "ISLR"), blend = TRUE)

# Blended feature plot
p1 <- FeaturePlot(seurat_combined, features = c("SOST", "ISLR"), blend = TRUE)

# Set a threshold for SOST expression
threshold <- 0.5

# Filter cells with SOST expression >= 0.5
seurat_combined_filtered <- subset(seurat_combined, subset = SOST >= threshold)

# Blended feature plot for the filtered data
p2 <- FeaturePlot(seurat_combined_filtered, features = c("SOST", "ISLR"), blend = TRUE)

# Save the plot with sufficient width and height
ggsave("feature_plot_threshold.png", plot = p2, width = 10, height = 3, dpi = 600)
ggsave("feature_plot.png", plot = p1, width = 10, height = 3, dpi = 600)

p1
p2
```


# Adjusted Parameters for FeaturePlot 
```{r}
# Extract expression data for the features
blend_data <- FetchData(seurat_combined, vars = c("ISLR", "EMILIN1"))

# Calculate a co-expression score for sorting
blend_data$coexpression_score <- blend_data$ISLR + blend_data$EMILIN1

# Add this co-expression score as a metadata column for plotting
seurat_combined$coexpression_score <- blend_data$coexpression_score

# Create the blended plot with sorted cells (co-expression highlighted)
p11 <- FeaturePlot(
  seurat_combined,
  features = c("ISLR", "EMILIN1"),, 
  min.cutoff = 0.0001, 
  max.cutoff = 5,
  blend = TRUE,
  order = TRUE  # Ensures cells with higher values are plotted on top
) +
  theme(legend.position = "bottom") +  # Adjust legend position
  guides(
    color = guide_legend(
      title = "Expression",
      title.theme = element_text(size = 8)
    )
  ) + 
  theme(legend.text = element_text(size = 8))

p11

# Save the plot
ggsave("feature_plot_adjust.png", plot = p11, width = 10, height = 3, dpi = 600)
```


# Which Cells in Specific Cell-Type Co-Express with Specific Gene of Interest?
## This code determines the number of cells that co-express with 2 genes (i.e. ISLR/RPL11)
```{r}
# Assuming 'seurat_combined' has metadata that indicates cell type in a column
smc_cells <- subset(seurat_combined, subset = orig.ident == "SMC")

# Ensure Gene is in the dataset
if (!"ISLR" %in% rownames(smc_cells)) {
  stop("Gene is not in the dataset.")
}

# Create a binary matrix: 1 if gene is expressed, 0 otherwise
binary_expression <- GetAssayData(smc_cells, assay = "RNA", layer = "data") > 0

# Select cells where Gene is expressed
islr_cells <- binary_expression["ISLR", ]

# Count co-expression for each gene
coexpression_counts <- rowSums(binary_expression[, islr_cells > 0])

# Remove Gene itself from the results
coexpression_counts <- coexpression_counts[rownames(binary_expression) != "ISLR"]
islr_co <- as.data.frame(coexpression_counts)

# Identify the top co-expressed gene
top_gene <- names(which.max(coexpression_counts))

# Display the top co-expressed gene and its count
list(
  top_gene = top_gene,
  coexpression_count = max(coexpression_counts)
)

# Optionally, store or visualize these genes
write.csv(islr_co, "islr_co_smc.csv")
```


# How Many Cells are Co-Expressing for Genes of Interest? (2 Only)
```{r}
# Fetch expression data
expression_data <- FetchData(seurat_combined, vars = c("CXCL6", "NT5E"))

# Set a threshold for expression (you can adjust this based on your dataset)
threshold <- 0.08  # For example, count cells expressing more than 0

# Identify co-expressing cells
co_expressed_cells <- sum(expression_data$CXCL6 > threshold & expression_data$NT5E > threshold)

# Output the count
print(co_expressed_cells)

```


# What is Average Expression for Specific Cell-Type?
```{r}
# 1. Subset Seurat object for the 'NeonatalCDC' cell type
neonatalCDC_cells <- subset(seurat_combined, idents = "NeonatalCDC")

# 2. Get the expression data for all genes (default assay is 'RNA')
expression_data_neonatalCDC <- GetAssayData(neonatalCDC_cells, assay = "RNA", slot = "data")

# 3. Calculate the average expression for each gene across the 'NeonatalCDC' cells
average_expression_neonatalCDC <- mean(expression_data_neonatalCDC)

# 4. Output the average expression for all genes in the 'NeonatalCDC' cell type
cat("Average expression in NeonatalCDC for all genes:", average_expression_neonatalCDC, "\n")


```


# Secretome Factors Analysis from hHiPC Proteomics
```{r}
secdf <- read.csv("Secretome_List_NT.csv")
secdf

# Extract gene names from the specified column
Gene <- secdf$Gene.Names # Replace 'gene' with the correct column name if different

secfiltered_df <- as.data.frame(Gene)
```


# What Is Average Expression for Specific Gene Across Cell Types?
```{r}
# Fetch expression data for SOST across all cells
sost_data <- FetchData(seurat_combined, vars = "SOST")

# Add the orig.ident (cell type) as a metadata column
sost_data$orig.ident <- seurat_combined$orig.ident

# Calculate average expression for SOST by cluster
sost_avg_by_cluster <- aggregate(sost_data$SOST, by = list(sost_data$orig.ident), FUN = mean)

# Rename the columns for clarity
colnames(sost_avg_by_cluster) <- c("Cluster", "Avg_Gene_Expression")

# Display the average expression levels
print(sost_avg_by_cluster)
```


# Which Genes Follow SOST Expression Profile in Single Cell Dataset?
```{r}
# Step 1: Calculate average expression levels by cluster
avg_expression <- AggregateExpression(seurat_combined, group.by = "orig.ident")

# Extract average expression data
avg_expression_data <- avg_expression$RNA  # Assuming the data is stored in the 'RNA' assay

# Check the row names (genes)
gene_names <- rownames(avg_expression_data)
print(gene_names[1:5])  # Display the first 5 gene names for a quick check

# Check the column names (cell types)
cell_types <- colnames(avg_expression_data)
print(cell_types)  # Display all column names to verify the cell types


# Define thresholds
threshold <- 0.009833558      # Set the threshold for expression in AdultCDC and NeonatalCDC
low_threshold <- 0.009833558 # Set the low threshold for expression in EC, SMC, and CF

# Filter genes based on expression levels
selected_genes_onlycdc <- gene_names[
  avg_expression_data[,"AdultCDC"] > threshold &
  avg_expression_data[,"NeonatalCDC"] > threshold &
  avg_expression_data[,"EC"] <= low_threshold &
  avg_expression_data[,"SMC"] <= low_threshold &
  avg_expression_data[,"CF"] <= low_threshold
]

# View selected genes
str(selected_genes_onlycdc)

# Convert selected genes to a data frame
selected_genes_df_onlycdc <- data.frame(Gene = selected_genes_onlycdc)

# Specify the file path where you want to save the CSV
file_path <- "selected_genes_onlycdc.csv"  # Change the path if needed
```

# Compare with hHiPC Proteomic Data (Single Cell Gene Expression Threshold Converted in 
# Previous Chunk)
```{r}
# Remove leading/trailing whitespace and convert to uppercase
selected_genes_df_onlycdc$Gene <- trimws(toupper(selected_genes_df_onlycdc$Gene))
secfiltered_df$Gene <- trimws(toupper(secfiltered_df$Gene))

# Assuming both data frames have a column named 'Gene'
common_genes_df <- merge(selected_genes_df_onlycdc, secfiltered_df, by = "Gene")

# View the new data frame with common genes
print(common_genes_df)
```

# Plot Common Proteomic and SCRNAseq Genes
```{r}
# Convert the Gene column in common_genes_df to a character vector
features <- as.character(common_genes_df$Gene)

# Create a violin plot for the common genes in the Seurat object
VlnPlot(seurat_combined, features = features)
```

# Now Add BMP9 Secretome Factors hHiPC
```{r}
secdf <- read.csv("Secretome_List_B9add.csv")

# Extract gene names from the specified column
Gene <- secdf$Gene.Names # Replace 'gene' with the correct column name if different

secfiltered_df_new <- as.data.frame(Gene)
```


# Which Genes Following SOST Expression Profile from BMP Dataset?
```{r}
# Step 1: Calculate average expression levels by cluster
avg_expression <- AggregateExpression(seurat_combined, group.by = "orig.ident")

# Extract average expression data
avg_expression_data <- avg_expression$RNA  # Assuming the data is stored in the 'RNA' assay

# Check the row names (genes)
gene_names <- rownames(avg_expression_data)
print(gene_names[1:5])  # Display the first 5 gene names for a quick check

# Check the column names (cell types)
cell_types <- colnames(avg_expression_data)
print(cell_types)  # Display all column names to verify the cell types


# Define thresholds
threshold <- 0      # Set the threshold for expression in AdultCDC and NeonatalCDC
low_threshold <- 0 # Set the low threshold for expression in EC, SMC, and CF

# Filter genes based on expression levels
selected_genes_onlycdc <- gene_names[
  avg_expression_data[,"AdultCDC"] > threshold &
  avg_expression_data[,"NeonatalCDC"] > threshold &
  avg_expression_data[,"EC"] <= low_threshold &
  avg_expression_data[,"SMC"] <= low_threshold &
  avg_expression_data[,"CF"] <= low_threshold
]


# View selected genes
str(selected_genes_onlycdc)

# Convert selected genes to a data frame
selected_genes_df_onlycdc <- data.frame(Gene = selected_genes_onlycdc)

# Specify the file path where you want to save the CSV
file_path <- "selected_genes_onlycdc.csv"  # Change the path if needed

# Remove leading/trailing whitespace and convert to uppercase
selected_genes_df_onlycdc$Gene <- trimws(toupper(selected_genes_df_onlycdc$Gene))
secfiltered_df_new$Gene <- trimws(toupper(secfiltered_df_new$Gene))

# Assuming both data frames have a column named 'Gene'
common_genes_df <- merge(selected_genes_df_onlycdc, secfiltered_df_new, by = "Gene")

# View the new data frame with common genes
print(common_genes_df)
```

# Plot Common Genes
```{r}
# Convert the Gene column in common_genes_df to a character vector
features <- as.character(common_genes_df$Gene)

# Create a violin plot for the common genes in the Seurat object
VlnPlot(seurat_combined, features = features)
```

# Save CSV
```{r}
# Specify the file path where you want to save the CSV
output_file_path <- "C:/Users/micha/OneDrive/Documents/SingleCell_CSphere/CSPhere/secfiltered_df.csv"

# Save the data frame as a CSV
write.csv(secfiltered_df, file = output_file_path, row.names = FALSE)
```


```{r}
save.image(file = "my_environment.RData")
```

```{r}
sessionInfo()
```

